The classical mutual information in mean-field spin glass models

### 2015-12-14 sk_model_v20.tex
15/12/2015

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[t]
\includegraphics*[width=0.93\linewidth]{./draft_figs/Renyi_MC_v2}
%\includegraphics*[width=0.93\linewidth]{./Renyi_MC_v2}
\caption{ The classical disorder-averaged R\'enyi entropy per spin $[S_2(\omega)]/N$  
 in the S-K model on the $2$-sheets booklet: 
 $[S_2(\omega)]/N$ as a function of the inverse temperature $\beta$. The different panels 
 correspond to different booklet aspect ratios $\omega=3/4,1/2,1/4$ (see Fig.~\ref{cartoon}). 
 The triangles, squares and rhombi denote the Monte Carlo results for 
 booklets with $N=32,64,128$ spins per sheet. The Monte Carlo error bars are often 
 smaller than the symbol size. The plus symbols are the numerical 
 extrapolations to the thermodynamic limit. The inset in panel (a) shows 
 $[S_2]/N$ at fixed $\beta=2$, plotted versus $N^{-2/3}$.The dotted line is 
 a linear fit. In all the panels  the dash-dotted lines are the  
 analytical results in the replica-symmetry (RS) approximation. 
 The circles are the analytical results obtained 
 using the one-step replica symmetry breaking (1-RSB) ansatz.
}
\label{Renyi-MC}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Here we discuss the R\'enyi entropies and the mutual information in the 
infinite-range Ising model withoud disorder defined by the hamiltonian 
%
\begin{equation}
\label{is-clean}
{\mathcal H}=-\frac{J}{N}\sum\limits_{i<j}S_iS_j+h\sum_i S_i.
\end{equation}
%
Here we consider the situation without external magnetic field, i.e., 
$h=0$ and fix $J=1$.

The calculation of the booklet partition function $Z(\omega,n,\beta)$ is 
done using the same techniques as in section~\ref{}. The result reads 
%
\begin{multline}
\label{clean-is-Z}
Z(\omega,n,\beta)=
\exp\Big(-\frac{\beta}{2}n\Big)\Big(2\pi\frac{\beta}{N}\Big)^{-\frac{n}{2}}
\int\prod_{\delta}dz_\delta\exp N\Big\{\\
-\sum_\delta\Big[\frac{z^2_\delta}{2\beta}
-(1-\omega)\log(2\cosh z_\delta)\Big]+\omega\log(2\cosh\sum_\delta z_\delta)
\Big\}.
\end{multline}
%
Using the saddle point approximation one obtains 
%
\begin{multline}
\label{clean-is-F}
\log Z(\omega,n,\beta)\approx
N\Big\{-\sum_\delta\Big[\frac{z^2_\delta}{2\beta}
-(1-\omega)\log(2\cosh z_\delta)\Big]\\
+\omega\log(2\cosh\sum_\delta z_\delta)\Big\}, 
\end{multline}
%
where we neglected subleading contributions $o(N)$. The parameters 
$\{z_\delta\}_{\delta=1}^n$ are solutions of the saddle point equations 
%
\begin{equation}
\label{clean-is-spe}
0=z_\delta-\beta(1-\omega)\tanh z_\delta-\beta\omega\tanh\big(\sum_\delta 
z_\delta\big),\quad \forall\delta.
\end{equation}
%
For $n=2$, Eq.~\eqref{clean-is-spe} become 
%
\begin{align}
0=z_1-\beta(1-\omega)\tanh z_1-\beta\omega\tanh\big(z_1+z_2\big)\\
0=z_2-\beta(1-\omega)\tanh z_2-\beta\omega\tanh\big(z_1+z_2\big). 
\end{align}
%
The system has the solution $z_1=z_2=0$ for $\beta\le\beta_c=1/(1+\omega)$, 
and $z_1=z_2\ne 0$ for $\beta>\beta_c$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/saddle_clean}
%\includegraphics*[width=0.9\linewidth]{./RSB0_saddle_ht}
\caption{
 The S-K model on the $2$-sheets booklet in the 
 paramagnetic phase: the solution $q_0$ of the saddle point 
 equation~\eqref{saddle-high-t} plotted as a function of the booklet aspect 
 ratio $\omega$ and several values of the inverse temperature $\beta=1/T$. 
 At $\omega=1$ one has $q_0=1,\forall\beta$. In the zero-temperature limit 
 one has $q_0\to 1,\forall\omega$. The straight line is the infinite 
 temperature result. 
}
\label{saddle_clean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/renyi_clean}
%\includegraphics*[width=0.9\linewidth]{./RSB0_saddle_ht}
\caption{
 The S-K model on the $2$-sheets booklet in the 
 paramagnetic phase: the solution $q_0$ of the saddle point 
 equation~\eqref{saddle-high-t} plotted as a function of the booklet aspect 
 ratio $\omega$ and several values of the inverse temperature $\beta=1/T$. 
 At $\omega=1$ one has $q_0=1,\forall\beta$. In the zero-temperature limit 
 one has $q_0\to 1,\forall\omega$. The straight line is the infinite 
 temperature result. 
}
\label{saddle_clean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\bibitem{alba-2013}
V.~Alba, J.\ Stat.\ Mech.\ (2013) P05013. 
\end{thebibliography}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[t]
\includegraphics*[width=0.93\linewidth]{./draft_figs/Renyi_MC_v2}
%\includegraphics*[width=0.93\linewidth]{./Renyi_MC_v2}
\caption{ The classical disorder-averaged R\'enyi entropy per spin $[S_2(\omega)]/N$  
 in the S-K model on the $2$-sheets booklet: 
 $[S_2(\omega)]/N$ as a function of the inverse temperature $\beta$. The different panels 
 correspond to different booklet aspect ratios $\omega=3/4,1/2,1/4$ (see Fig.~\ref{cartoon}). 
 The triangles, squares and rhombi denote the Monte Carlo results for 
 booklets with $N=32,64,128$ spins per sheet. The Monte Carlo error bars are often 
 smaller than the symbol size. The plus symbols are the numerical 
 extrapolations to the thermodynamic limit. The inset in panel (a) shows 
 $[S_2]/N$ at fixed $\beta=2$, plotted versus $N^{-2/3}$.The dotted line is 
 a linear fit. In all the panels  the dash-dotted lines are the  
 analytical results in the replica-symmetry (RS) approximation. 
 The circles are the analytical results obtained  using the one-step replica 
 symmetry breaking (1-RSB) ansatz. In panel (c) the dotted line (clean Ising) is the 
 analytic result for the infinite-range Ising model without disorder. 
}
\label{Renyi-MC}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Here we discuss the R\'enyi entropy $S_2$ and the associated mutual information 
${\mathcal I}_2$ in the infinite-range Ising model without disorder defined by 
the hamiltonian 
%
\begin{equation}
\label{is-clean}
{\mathcal H}=-\frac{J}{N}\sum\limits_{i<j}S_iS_j+h\sum_i S_i.
\end{equation}
%
We consider the situation without external magnetic field, i.e., $h=0$ and 
fix $J=1$.

%##########################################################################
\subsection{Saddle point approximation}

The calculation of the booklet partition function $Z(\omega,n,\beta)$ 
(see section~\ref{booklet}) can be done using the same techniques as in 
section~\ref{the-model}. The result reads 
%
\begin{multline}
\label{clean-is-Z}
Z(\omega,n,\beta)=
\exp\Big(-\frac{\beta}{2}n\Big)\Big(2\pi\frac{\beta}{N}\Big)^{-\frac{n}{2}}
\int\prod_{\delta}dz_\delta\exp N\Big\{\\
-\sum_\delta\Big[\frac{z^2_\delta}{2\beta}
-(1-\omega)\log(2\cosh z_\delta)\Big]+\omega\log(2\cosh\sum_\delta z_\delta)
\Big\}.
\end{multline}
%
Using the saddle point approximation, from~\eqref{clean-is-Z} one obtains 
$F(\omega,n,\beta)\equiv\log Z(\omega,n,\beta)$ as 
%
\begin{multline}
\label{clean-is-F}
F(\omega,n,\beta)\approx
N\Big\{-\sum_\delta\Big[\frac{z^2_\delta}{2\beta}
-(1-\omega)\log(2\cosh z_\delta)\Big]\\
+\omega\log(2\cosh\sum_\delta z_\delta)\Big\}, 
\end{multline}
%
where we neglected subleading contributions $o(N)$. The parameters 
$\{z_\delta\}_{\delta=1}^n$ are solutions of the saddle point equations 
%
\begin{equation}
\label{clean-is-spe}
0=z_\delta-\beta(1-\omega)\tanh z_\delta-\beta\omega\tanh\big(\sum_\delta 
z_\delta\big),\quad \forall\delta.
\end{equation}
%
For $n=2$, the system~\eqref{clean-is-spe} becomes 
%
\begin{align}
& 0=z_1-\beta(1-\omega)\tanh z_1-\beta\omega\tanh\big(z_1+z_2\big)\\
& 0=z_2-\beta(1-\omega)\tanh z_2-\beta\omega\tanh\big(z_1+z_2\big). 
\end{align}
%
The system has the solutions $z_1=z_2=z_s=0$ for $\beta\le\beta_c=1/(1+\omega)$, 
and $z_1=z_2=z_s\ne 0$ for $\beta>\beta_c$. Note that $\beta_c$ is different 
in presence of disorder (compare with~\eqref{tc}).

The solution $z_s$ of the saddle point equations~\eqref{clean-is-spe} is shown 
in Fig.~\ref{saddle_clean} for $\omega=0,1/2,1$. The vertical-dotted line 
is $\beta_c$ for $\omega=1/2$. Notice that at low temperatures $z_s$ is 
independent on $\omega$, and one has $z_s\sim\beta$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/saddle_clean}
%\includegraphics*[width=0.9\linewidth]{./saddle_clean}
\caption{ The infinite-range Ising model on the $2$-sheets booklet: saddle point 
 approximation. The solution $z_s$ of the saddle point equations~\eqref{clean-is-spe} 
 plotted as a function of the inverse temperature $\beta$ and for booklet aspect 
 ratio $\omega=0,1/2,1$. The vertical dotted line denotes the critical inverse 
 temperature for $\omega=1/2$. 
}
\label{saddle_clean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%##########################################################################
\subsection{R\'enyi entropy and mutual information: saddle point results}

It is straightforward to obtain the R\'enyi entropies $S_n=(F(\omega=0,n,\beta)-
F(\omega,n,\beta))/(n-1)$ using~\eqref{clean-is-F} and~\eqref{renyi}. 
Figure~\ref{renyi_clean} plots the entropy density $S_2/N$ versus the inverse 
temperature $\beta$. $S_2$ exhibits a volume-law behavior $S_2\propto N$ for 
all values of $\beta$ and $\omega$. The different curves correspond to different 
booklet aspect ratios $\omega$. Clearly, $S_2$ exhibits the volume-law behavior 
$S_2\propto N$ at all temperatures. For $\beta\le\beta_c$ one has the flat behavior 
$S_2=\omega\log(2)$, whereas $S_2$ is vanishing in the low-temperature regime. 
The full symbol (rhombi) in Figure~\ref{renyi_clean} are exact results for a 
booklet with $N=400$ spins per sheet (see next section), and are in good 
agreement with the saddle point approximation results.

The mutual information per spin ${\mathcal I}_2/N$ is reported in 
Figure~\ref{mi_clean}. The lines are the analytical results for booklet aspect 
ratios $\omega=1/20,1/4,1/2$, while the full symbols are the exact results for 
the booklet with $N=400$ spins per sheet. Clearly, ${\mathcal I}_2/N$ is exactly zero 
in the high-temperature region for $\beta\le1/2$, it exhibits at maximum 
at $\beta\approx 1$, and it is vanishing in the limit $\beta\to\infty$. 
Similar to Figure~\ref{reny_clean} the full symbols are exact results for 
a booklet with $N=400$ spins per sheet.

Notice that similar features are observed for the S-K model (see 
section~\ref{I2-section}). In particular, one should observe that ${\mathcal I}_2/N$ 
does not exhibit any crossing at the critical point $\beta_c$. However, the information 
about critical behavior is encoded in the finite-size corrections of ${\mathcal I}_2/N$ 
(see next section), which are neglected in the saddle point treatment. 
Finally, one should also stress that a dramatically different behavior is observed 
in the mutual information obtained from the von Neumann entropy ${\mathcal I}\equiv 
\lim_{n\to 1}{\mathcal I}_n$. Specifically, no volume law is present for ${\mathcal I}$, 
which, instead, exhibits a logarithmic divergence as ${\mathcal I}=1/4\log(N)$ at the 
critical point $\beta_c=1$ (see Ref.~\onlinecite{wilms-2012}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/renyi_clean}
%\includegraphics*[width=0.9\linewidth]{./renyi_clean}
\caption{ The infinite-range Ising model on the $2$-sheets booklet in the thermodynamic 
 limit: the classical R\'enyi entropy density $S_2/N$, with $N$ the total number of spins 
 on a single sheet of the booklet. $S_2/N$ is plotted against the inverse temperature 
 $\beta$ and for booklet aspect ratio $\omega=1/4,1/2,1$. Note tht at high temperature 
 $S_2=\omega\log(2)$. The full symbols denote the exact numerical results for a finite 
 system with $N=400$. 
}
\label{renyi_clean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/mi_clean}
%\includegraphics*[width=0.9\linewidth]{./renyi_clean}
\caption{ The infinite-range Ising model on the $2$-sheets booklet in the thermodynamic 
 limit: the classical mutual information density ${\mathcal I}_2/N$ plotted against the 
 inverse temperature $\beta$ and for booklet aspect ratio $\omega=1/4,1/2,1$. Note that  
 ${\mathcal I}_2=0$ for $\beta\le 1/2$. The full symbols are the the exact numerical 
 results for a finite system with $N=400$. 
}
\label{mi_clean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%##########################################################################
\subsection{R\'eny entropy and mutual information: Exact treatment}

For finite number of spins $N$, $S_2$ and ${\mathcal I}_2$ can be calculated 
exactly using the results obtained in Ref.~\onlinecite{wilms-2012}. The eigenstates 
of~\eqref{is-clean} are product states. They can be characterized as $|p,i\rangle$, 
with $p$ the total number of up spins (the remaining $N-p$ being down) and $i$ 
labelling the $B(N,p)\equiv N!/(p!(N-p)!)$ eigenstates with the same $p$.

The partition function of the model on a single sheet is given as 
%
\begin{equation}
Z(\beta)\equiv\textrm{Tr}\rho=\sum\limits_{p=0}^NB(N,p)e^{N\beta(\frac{p}{N}-
\frac{1}{2})^2},
\end{equation}
%
where $\rho\equiv 1/Ze^{-\beta {\mathcal H}}$ is the thermal density matrix. 
Given a bipartition of the spins into two groups $A$ and $B$ containing $N_A$ 
and $N-N_A$ spins, respectively, the reduced density matrix for $A$ is obtained 
as~\cite{wilms-2012} 
%
\begin{equation}
\rho_A\equiv\frac{1}{Z}\textrm{Tr}_B\sum\limits_{p,i}e^{N(\beta\frac{p}{N}-
\frac{1}{2})^2}|p,i\rangle\langle p,i|. 
\end{equation}
%
By performing the trace over part $B$ one obtains 
%
\begin{equation}
\rho_A=\sum\limits_{p_A,i_A}R(p_A)|p_A,i_A\rangle\langle p_A,i_A|,
\end{equation}
%
where $|p_A,i_A\rangle$ form a basis for part $A$ of the system and 
%
\begin{equation}
R(p_A)\equiv\frac{1}{Z}\sum\limits_{p_B=0}^{N-N_A}B(N-N_A,p_B)e^{N\beta
(\frac{p_A}{N}+\frac{p_B}{N}-\frac{1}{2})^2}.
\end{equation}
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/sub_leading}
%\includegraphics*[width=0.9\linewidth]{./renyi_clean}
\caption{ The infinite-range Ising model on the $2$-sheets booklet: saddle point corrections. 
 The correction for the mutual information ${\mathcal I}_2-{\mathcal I}_2^{(sp)}$ are 
 plotted against the inverse temperature $\beta$. The data are for a booklet with aspect 
 ratio $\omega=1/2$. Here ${\mathcal I}_2^{(sp)}$ are the same data as in Figure~\ref{mi_clean} 
 (dotted line). The vertical lines denote the critical temperatures $\beta_c$ for $\omega=1,
 1/2,0$ (from left to right). The horizontal dash-dotted line is $\log(2)$. Note the 
 (logarithmically) divergent behavior at $\beta_c$. 
}
\label{sp_corr}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/clean_log}
%\includegraphics*[width=0.9\linewidth]{./renyi_clean}
\caption{ The infinite-range Ising model on the $2$-sheets booklet: 
 Scaling of the corrections of the mutual information ${\mathcal I}_2$ 
 at the booklet critical points $\beta_c=1/2,2/3,1$ (see 
 vertical lines in Figure~\ref{sp_corr}). We plot $|{\mathcal I}_2-
 {\mathcal I}_2^{(sp)}|$ versus $\log(N)$, with $N$ the number of spins 
 on one sheet of the booklet. Here ${\mathcal I}_2^{(sp)}$ is the same 
 as in Figure~\ref{mi_clean} (dotted line).  
}
\label{clean_log}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\bibitem{alba-2013}
V.~Alba, J.\ Stat.\ Mech.\ (2013) P05013.

\bibitem{wilms-2012}
J.~Wilms, J.~Vidal, F.~Verstraete, and S.~Dusuel, J.\ Stat.\ Mech.\ (2012) P01023.

\end{thebibliography}
